---
title: "Derivatives and Partial Derivatives with grad"
author: "W. Bauer"
date: "03/03/2015"
output: html_document
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
# load package numDeriv, installing if necessary
if(!require(numDeriv)){
  install.packages("numDeriv")
  library(numDeriv)
  }
source("../R/utilities.R")
```

1. The point of this unit is to introduce derivatives and partial derivatives in a general setting, and to introduce numerical differentiation with `grad`.
    + Previously, we noted the simplest example of a derivative is the slope of a line.
    + We also noted that the closer you look at a smooth curve, the more it looks like a line. This allows us to assign a slope at any point on a smooth curve by taking a limit.
    ```{r fig.show='hold', fig.align='center', echo=FALSE}
makegrid(0:4, ylab="f(x)", major_axes=FALSE, main="Derivative of a function of 1 variable", cex.lab=2)
local({
  x <- seq(0,4,by=.1)
  y <- 2*sin(pi*x/4)
  lines(x, y, lwd=3)
  y1 <- 2*sin(pi*1.5/4)
  slope <- (pi/2)*cos(pi*1.5/4)
  y2 <- y1 + slope
  points(1.5, y1, pch=19, col="blue", cex=2.0)
  segments(.5, y1-slope, 2.5, y2, lwd=3, col="blue")
  text(.25, 3.0, expression(over(d*f(x), d*x) %==% lim(over(f(x + Delta*x) - f(x), Delta*x), Delta*x %->% 0)), pos=4)
  })
```

2. Numerical derivatives with `grad`
    + Derivatives of many functions can be found analytically. The figure shows how to find the derivative of $x^2$ using only the definition and elementary algebra.
    ```{r fig.show='hold', fig.align='center', echo=FALSE}
plot(0:1, 0:1, type='n', xlab="", ylab="", xaxt="n", yaxt="n", bty="n", main=expression(paste("The analytic derivation of ", over(d*x^2,d*x))))
cr <- .2
text(0, .9, expression(over(d*x^2,d*x) %==% lim(over((x+Delta*x)^2-x^2, Delta*x), Delta*x %->% 0)), pos=4)
text(0, .9-cr, expression(paste(over(d*x^2,d*x) == lim(over(x^2+2*x*Delta*x + Delta*x^2-x^2, Delta*x), Delta*x %->% 0), "    ( expanding ", (x+Delta*x)^2, " )")), pos=4)
text(0, .9-2*cr, expression(paste(over(d*x^2,d*x) == lim(over(2*x*Delta*x + Delta*x^2, Delta*x), Delta*x %->% 0),"  ( cancelling ", x^2 - x^2, " )")), pos=4)
text(0, .9-3*cr, expression(paste(over(d*x^2,d*x) == lim(2*x + Delta*x, Delta*x %->% 0), "    ( dividing out ", Delta*x, " )")), pos=4)
text(0, .9-4*cr, expression(over(d*x^2, d*x) == 2*x.), pos=4)
```
    + Here, however, we will focus on estimating derivatives numerically, using the function, `grad`, in package numDiff. (The name is a short form of "gradient," a term which will be explained later.)
    + `grad` takes two arguments, a function and a point at which the derivative of the function is to be estimated.
    + For instance `grad(function(x)x^2, 1)` will estmate the derivative of `x^2` at x=1. You know from the analytic derivation that this should be 2.
    + In the example, grad's first argument, `function(x)x^2` is an R function which takes one numerical argument and returns the value `x^2` at the point `x` represented by that argument.
    + Using common functions such as cos, sin, exp, log, 1/x, have students compare their numerical and analytic derivatives. The follow code illustrates how that might be done. Note that `grad` is vectorized for functions of 1 variable but has nothing analogous for functions of several. I think that is confusing, but vectorization comes in handy here.
    ```{r fig.show='hold', fig.align='center'}
# Show that the derivative of sin is cos.
x <- seq(0, 2*pi, by=.2)
# swirl does this
plot(x, cos(x), type='l', lwd=2, xlab="x", ylab="sin(x)", main=expression(over(d, d*x)~~sin(x) == cos(x)))
abline(h=0)
legend('bottomleft', "cos(x)", lwd=2, cex=1.5)
legend('bottomright', expression(over(d, d*x)~~sin(x)), pch=19, col="blue", cex=1.5)
# student does this
deriv <- grad(sin, x)
points(x, deriv, pch=19, col="blue", cex=1.5)
```

3. Gradients on Maunga Whau
    + In general, we may have functions of several variables, such as this topographic map of Maunga Whau volcano. Here f(x, y) is altitude, x is northward distance from a reference point, and y is westward distance from the reference point.
    ```{r fig.show='hold', fig.align='center', echo=FALSE}
  source("../R/maungaWhau.R")
  bg <- par("bg")
  par(bg = "lightblue")
  res <- persp(x, y, z, theta = 120, phi = 15, col = fill, scale = FALSE, xlab="x (South to North)", ylab="y (East to West)", zlab = "f(x, y)")
  title(main="Maunga Whau\nOne of 50 Volcanoes in the Auckland Region.", font.main = 4)
  par(bg=bg)
```
    + For this case there are *partial derivatives* in which estimate the slope with respect to one variable while holding the others constant.
    ```{r fig.show='hold', fig.align='center', echo=FALSE}
plot(0:1, 0:1, type='n', xlab="", ylab="", main="Partial Derivatives", xaxt='n', yaxt='n', bty='n')
text(0, 0.8, expression(over(partialdiff*f(x,y), partialdiff*x) %==% lim(over(f(x + Delta*x, y) - f(x, y), Delta*x), Delta*x %->% 0)), pos=4, cex=1.5)
text(0, 0.2, expression(over(partialdiff*f(x,y), partialdiff*y) %==% lim(over(f(x, y + Delta*y) - f(x, y), Delta*y), Delta*y %->% 0)), pos=4, cex=1.5)
```
    + At a point on Maunga Whau, the partial derivatives represent slopes in the x and y directions. In the case pictured, the partial derivatives are indicated by the slopes of the lines labeled x and y.
    ```{r fig.show='hold', fig.align='center', echo=FALSE}
bg <- par("bg")
par(bg="lightblue")
res <- persp(x, y, z, theta = 120, phi = 15, col = fill, scale = FALSE, box=FALSE, main="Partial Derivatives")
par(bg=bg)
i <- 89-23# 89-16
j <- 63-18 # 63-32
myloc <- trans3d(x[i], y[j], z[i, j], res)
points(myloc$x, myloc$y, pch=19, cex=1.5, col="yellow")
# Create a line segment through v[1], v[2], z[i,j] in the x direction with appropriate slope.
a <- trans3d(x[i+5], y[j], z[i+5, j], res)
b <- trans3d(x[i-5], y[j], z[i-5, j], res)
text(b$x, b$y, "x", cex=1.5, col="yellow", pos=4)
segments(a$x, a$y, b$x, b$y, col="yellow", lwd=3)
a <- trans3d(x[i], y[j+5], z[i, j+5], res)
b <- trans3d(x[i], y[j-5], z[i, j-5], res)
text(b$x, b$y, "y", cex=1.5, col="yellow", pos=2)
segments(a$x, a$y, b$x, b$y, col="yellow", lwd=3)

```
    + The *gradient* of a function of more than one variable is a vector of its partial deriviatives. It is usually symbolized by a nabla (upside down triangle,) as illustrated in the figure.
    ```{r fig.show='hold', fig.align='center', echo=FALSE}
plot(0:1, 0:1, type='n', xlab="", ylab="", main="Gradient of f(x,y)", xaxt='n', yaxt='n', bty='n')
text(.5, .5, expression(nabla*f(x, y) %==% bgroup("(",atop(over(partialdiff*f(x,y), partialdiff*x), over(partialdiff*f(x,y), partialdiff*y)),")")), cex=2)
```
    + It can be shown that the gradient indicates the magnitude and direction of steepest ascent. The gradient at a point on Maunga Whau is shown here.
    ```{r fig.show='hold', fig.align='center', echo=FALSE}
bg <- par("bg")
par(bg="lightblue")
res <- persp(x, y, z, theta = 120, phi = 15, col = fill, scale = FALSE, box=FALSE, main="Gradient at a Point on Maunga Whau")
par(bg=bg)
points(myloc$x, myloc$y, pch=19, cex=1.5, col="yellow")
g <- grad(mw, c(x[i], y[j]))
g <- 50*g/sqrt(sum(g^2))
v <- g + c(x[i], y[j])
i1 <- round(v[1]/10)
j1 <- round(v[2]/10)
tip <- trans3d(v[1], v[2], z[i1, j1], res)
arrows(myloc$x, myloc$y, tip$x, tip$y, col="yellow", lwd=3)
```
    + Like derivatives, gradients can be estimated using the function, `grad`, in package numDiff.
    + As before, `grad` takes two arguments, a function and a point at which the gradient of the function is to be estimated. But in this case, the "point" is a vector.
    + If the function depends on two variables, the vector has length two, if three length three, and so on.
    + We've written `mw(point)` which gives the altitude of Maunga Whau as a function of point coordinates. (Actually it gives twice the altitude, since that's the way z is calibrated.)
    + Illustrate the use of `mw` by finding the altitude at the pictured points.
    ```{r fig.show='hold', fig.align='center', echo=FALSE}
place <- function(i, j, res, col="yellow"){
  temp <- trans3d(x[i], y[j], z[i,j], res)
  points(temp$x, temp$y, pch=19, cex=1.5, col=col)
  text(temp$x, temp$y, paste0("c(",x[i],",",y[j],")"), pos=3, col=col, lwd=2)
}
bg <- par("bg")
par(bg="lightblue")
res <- persp(x, y, z, theta = 120, phi = 15, col = fill, scale = FALSE, box=FALSE, main="Points and Gradients on Maunga Whau")
par(bg=bg)
place(68,32,res)
place(59,40,res)
place(37,47,res)
```
    
    ```{r}
mw(c(670, 310))
mw(c(580, 390))
mw(c(360, 460))
```
    + Illustrate `grad` by finding the gradient at those same points
    ```{r}
grad(mw,c(670, 310))
grad(mw,c(580, 390))
grad(mw,c(360, 460))
```
    + The gradient at `c(670, 310)` is zero because that location is a *local maximum*, i.e., the top of a hill. Note that it is not the highest point on the map, hence is called local.

4. Gradient ascent
    + Ascend Maunga Whau by gradient ascent. The climber goes 50 meters in the gradient, are steepest ascent, direction.
    + This could be a demo or an interactive question where the user gets to guess which of 3 starting points will ascend to the rim of the volcano.
    ```{r fig.show='hold', fig.align='center', echo=FALSE}
bg <- par("bg")
par(bg="lightblue")
res <- persp(x, y, z, theta = 120, phi = 15, col = fill, scale = FALSE, box=FALSE, main="Climbing Maunga Whau")
par(bg=bg)
plunk <- function(v, col="yellow"){
  i <- ceiling(1e-10+v[1]/10)
  j <- ceiling(1e-10+v[2]/10)
  temp <- trans3d(x[i], y[j], z[i,j], res)
  points(temp$x, temp$y, pch=19, cex=1.25, col=col)
  c(x[i], y[j])
}
step <- function(v){
  if(sum(is.na(mw(v))))return(v)
  g <- grad(mw, v)
  gnorm <- sqrt(sum(g^2))
  if(gnorm > 0)g <- 50*g/gnorm
  v <- v + g
  plunk(v)
  v
}
# Choose a starting point and watch the hiker climb. These do vaguely interesting things:
# Run as real time "movies", they are cute. 
# v <- c(520, 550) 
# v <- c(800, 400)
# v <- c(560, 520)

ascend <- function(v, steps=10){
  plunk(v, col="orange")
  for(i in 1:10){
    Sys.sleep(0.25)
    v <- step(v)
  }
}

ascend(c(520, 550))
ascend(c(800, 400))
ascend(c(560, 520))
```
