---
title: "Lines, Curves, Tangents, Distance, Length, and Area"
author: "W. Bauer"
date: "01/10/2015"
output: html_document
---

1. The point of this unit (lesson or lessons) is to illustrate the main concepts visually, using plane figures.

2. Plane geometry is about lines and curves. Thanks to René Descartes, 1596-1650, it is also about Cartesian coordinates, which is what the light blue grid lines are about. 
  * We can specify any point in the plane by its x and y coordinates. The circle's center, for instance, has x coordinate 8 and y coordinate 6. One end of the line is at x=1 and y=6, the other at x=5, y=9. 
  * Using `points(x, y, pch=19, cex=3, col="green")` put a fat, green point at coordinates, x and y, of your choice; just be sure both x and y are between 0 and 10 so your point appears on the graph.
  * Thus, any point in the plane can be represented by a vector of 2 numbers. In R we'd represent the circle's center as `c(8,6)` and the endpoints of the line as `c(1,6)` and `c(5, )`.

```{r echo=FALSE, fig.show='hold'}
plot(c(0,10), c(0,10), type='n', xlab="x", ylab="y", main="Lines and curves")
abline(h=0:10, col="lightblue", lwd=2)
abline(v=0:10, col="lightblue", lwd=2)
segments(1,6,5,9, lwd=3)
points(c(1,5),c(6,9), pch=19, cex=1.5)
points(5, 9, pch=19, cex=1.5)
points(8,6,cex=25,lwd=3)
points(8,6, pch=19, cex=1.5, col=2)
x <- seq(-1, 11, by=.5)
lines(x, 1+sin(.2*pi*x), lwd=3)
```

3. A line has a length and a slope. Thanks to Pythagoras, we can calculate the length of a line segment

```{r echo=FALSE, fig.show='hold'}
plot(c(0,10), c(0,10), type='n', xlab="x", ylab="y", main="Line segments have slopes and lengths.")
abline(h=0:10, col="lightblue", lwd=2)
abline(v=0:10, col="lightblue", lwd=2)
segments(1,6,5,9, lwd=3)
points(c(1,5),c(6,9), pch=19, cex=1.5)
segments(1,6,5,6, lty=2, lwd=3, col=2)
segments(5,6,5,9, lty=2, lwd=3, col=2)
text(3.25, 5.5, expression(4), lwd=3, cex=1.5, col=2)
text(5.25, 7.5, expression(3), lwd=3, cex=1.5, col=2)
text(2.5, 8.5, expression(sqrt(4^2 + 3^2)), lwd=3, cex=1.5, col=1)
text(1.1, 8.5, "length = ", lwd=3, cex=1.5, col=1)
text(5.75, 5, expression(frac(3,4)), lwd=3, cex=1.5, col=2)
text(4.8, 5, "slope = ", lwd=3, cex=1.5, col=2)
```

3. The closer you look at a smooth curve, the more it looks like a line.