---
title: "Lines, Curves, Tangents, Distance, Length, and Area"
author: "W. Bauer"
date: "01/10/2015"
output: html_document
---

1. The point of this unit (lesson or lessons) is to illustrate the main concepts visually, using plane figures.

2. Plane geometry is about lines and curves. Thanks to René Descartes, 1596-1650, it is also about Cartesian coordinates, which is what the light blue grid lines are about. 
  * We can specify any point in the plane by its x and y coordinates. The circle's center, for instance, has x coordinate 8 and y coordinate 6. One end of the line is at x=1 and y=6, the other at x=5, y=9. 
  * Using `points(x, y, pch=19, cex=3, col="green")` put a fat, green point at coordinates, x and y, of your choice; just be sure both x and y are between 0 and 10 so your point appears on the graph.
  * Thus, any point in the plane can be represented by a vector of 2 numbers. In R we'd represent the circle's center as `c(8,6)` and the endpoints of the line as `c(1,6)` and `c(5, )`.

```{r echo=FALSE, fig.show='hold'}
plot(c(0,10), c(0,10), type='n', xlab="x", ylab="y", main="Lines and curves")
abline(h=0:10, col="lightblue", lwd=2)
abline(v=0:10, col="lightblue", lwd=2)
segments(1,6,5,9, lwd=3)
points(c(1,5),c(6,9), pch=19, cex=1.5)
points(5, 9, pch=19, cex=1.5)
points(8,6,cex=25,lwd=3)
points(8,6, pch=19, cex=1.5, col=2)
x <- seq(-1, 11, by=.5)
lines(x, 1+sin(.2*pi*x), lwd=3)
```

3. A line has a slope and a length. 
  * A line's slope is the simplest example of a *derivative*. In the figure, if x increases by 1, then y increases by 3/4. We say that 3/4 is the rate of change of y with respect to x, or the derivative of y with respect to x.
  * According to Pythagoras, "The square on the hypotenuse, is equal to the sum of the squares on its other two sides." Thus we can calculate the length of a line from the coordinates of its endpoints.
  * The square on the hypotenuse, $3^2+4^2$ in the figure, the simplest example of a *quadratic form*. It is called *quadratic* because of the squares, and *form*, is short for formula.
  * The line's length, of course, is the square root of the quadratic form.

```{r echo=FALSE, fig.show='hold'}
plot(c(0,10), c(0,10), type='n', xlab="x", ylab="y", main="Line segments have slopes and lengths.")
abline(h=0:10, col="lightblue", lwd=2)
abline(v=0:10, col="lightblue", lwd=2)
segments(1,6,5,9, lwd=3)
points(c(1,5),c(6,9), pch=19, cex=1.5)
segments(1,6,5,6, lty=2, lwd=3, col=2)
segments(5,6,5,9, lty=2, lwd=3, col=2)
text(3.25, 5.5, expression(4), lwd=3, cex=1.5, col=2)
text(5.25, 7.5, expression(3), lwd=3, cex=1.5, col=2)
text(2.5, 8.5, expression(sqrt(4^2 + 3^2)), lwd=3, cex=1.5, col=1)
text(1.1, 8.5, "length = ", lwd=3, cex=1.5, col=1)
text(5.75, 5, expression(frac(3,4)), lwd=3, cex=1.5, col=2)
text(4.8, 5, "slope = ", lwd=3, cex=1.5, col=2)
```

3. The closer you look at a smooth curve, the more it looks like a line.
  * In the figure, we're zooming in on the x axis around the coordinate x=1.
  * In the upper left we're looking at the interval between x=0 and x=2. In the upper right we zoom in to an interval 5 times smaller, and so on. By the time we're at an interval 50 times smaller, the line and the curve are indistiguishable by eye.
  * By a process such at this, we can assign a slope to a point on a curve. Such a slope is called the derivative of the curve and written $\frac{dy}{dx}$.

```{r echo=FALSE, fig.show='hold'}
local({
  mags <- c(1, 5, 10, 50)
   par(mfrow=c(2, 2))
 for(i in mags){
    x0 <- 1
    y0 <- x0^2
    x <- seq(x0 - 1/i, x0 + 1/i, length.out=50)
    y <- x^2
    plot(x, y, type='n', xlab=expression(x), ylab=expression(y), 
         main=paste0("Zoom = ", i*100,"%"))
    lines(x, y, lwd=3)
    points(x0, y0, pch=19, cex=1.5, col="red")
    abline(x0^2-2*x0*x0, 2*x0, lwd=3, lty=2, col="red")
  }
 par(mfrow=c(1,1))
})

```

